---
title: "Exploring the data sets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exploring the data sets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=5,
  fig.height=7
)
```

# Introduction and quick start

`flyer` is available on GitHub, and should make its way to CRAN eventually.
It can be installed via either of the following commands.

```{r install package, eval = F}
install.packages('devtools')
devtools::install_github('sagesteppe/flyer')

# remotes is very similar and a good alternative for this use case.
install.packages('remotes') 
remotes::install_github('sagesteppe/flyer')
```

To explore the data we'll load a couple of packages for handling it (`sf`, `dplyr`), plus a slew of packages for plotting with `ggplot2` (`ggnewscale`, `ggrepel`, `ggspatial`).
It might seem onerous to install all of these, but I bet once you see what they do you'll be quite excited to have them.

```{r setup, warning=FALSE, message=FALSE}
library(flyer)
library(dplyr) # for general data handling
library(sf) # for spatial data

library(ggplot2) # all for plotting the data 
library(ggnewscale) # for mapping multiple variables to an aesthetic. 
library(ggrepel) # for text based labels which move to minimize overlaps. 
library(ggspatial) # compasses and scale bars. 
```

We'll modify the number of graticules right off the bat. Note that the `pretty` function does not always return the requested `n`, so we wrap it in a small helper.

```{r define a carto ggplot theme}
graticNo <- function(polygon, nx, ny){
  
  bb <- round(st_bbox(polygon), 1)
  if(all(missing(nx) & missing(ny))) {nx <- 5;ny <- 5} else {
    if(all(missing(nx) & ! missing(ny))) {nx <- ny} else {
      if(all(missing(ny) & ! missing(nx))) {ny <- nx}
    } 
  } 

  xbreaks <- pretty(seq(bb[1], bb[3], length.out = nx), nx)
  ybreaks <- pretty(seq(bb[2], bb[4], length.out = ny), ny)
  
  return(
    list(
      x = xbreaks, y = ybreaks
    )
  )
}

brks <- graticNo(polygon = places, nx = 4, ny = 4)

theme_nautical <- function() {
  theme(
    
    aspect.ratio = 4/3,
    text = element_text(family = "Optima"),
    axis.title = element_text(colour = "#222823"),
    axis.text = element_text(colour = "#575A5E", face = "italic"),
    axis.text.x = element_text(hjust= 1, angle=45),
    axis.ticks = element_line(colour = "#575A5E"),
    
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 1, face = 'italic'),
    plot.caption = element_text(color = '#575A5E', face = 'italic'),
    
    panel.background = element_rect(fill = "#F4F7F5"),
    panel.border = element_rect(colour = NA, fill = NA),
    panel.grid.major = element_line(colour = "#A7A2A9", linetype = 'dotted', linewidth = 0.25),
    panel.grid.minor = element_blank(),
    
    legend.background = element_rect(fill = '#FEF9F3', color = '#A7A2A9'),
    legend.text = element_text(size = 5),
    legend.title = element_text(size = 7, hjust = 0.5), 
    legend.title.position = 'top',
    legend.key.size = unit(1,"line"),
    legend.position = "bottom",
    legend.spacing = unit(0.1, "line")
  )
}

bb <-st_bbox(
    c(xmin = -121, xmax = -105, ymin = 19, ymax = 34.5),  crs = st_crs(4326)
)

```

# The data sets

The places visited by the collectors can be loaded via `places`, and the route they took via `route`.
These are really the whole point of the package — and, spoiler alert, they're very simple!

But before we pull up `places` and `route`, let's pull up the `land` data set so we have some context to plot them on.


### land

We can read in some polygons depicting land from Natural Earth, via the `rnaturalearth` package. I love this package's functionality, even if I get real forgetful of their API calls (theme argument?).

```{r add land}
head(land)

ggplot() + 
  geom_sf(data = land, fill = '#F8F6F0') + 
  theme_nautical() +
  labs(title = 'land') + 
  coord_sf(xlim = c(bb[1], bb[3]), ylim = c(bb[2], bb[4])) 
```

For playing around with the data today, I don't want the different countries drawn separately, so we can union them.

```{r union land}
land <- st_union(land)

m <- ggplot() + 
  geom_sf(data = land, fill = '#F8F6F0') + 
  theme_nautical() + 
  labs(title = 'st_union(land)') + 
  coord_sf(xlim = c(bb[1], bb[3]), ylim = c(bb[2], bb[4])) 

m
```

### places

```{r plot the places}
data(places)
head(places)

m <- m + 
  geom_sf(data = places)  +
  labs(title = 'places', subtitle = '+ land') + 
  coord_sf(xlim = c(bb[1], bb[3]), ylim = c(bb[2], bb[4])) 
m
```

The places seem like they'll be better treated as text labels — we can apply them with `ggrepel::geom_text_repel`, which will nudge them to avoid conflicts with other plot elements.

```{r plot places as text, warning=F}
m <- m + 
  coord_sf(xlim = c(-118, -106), ylim = c(22, 31)) +
  ggrepel::geom_text_repel(
    data = places,
    aes(label = location_espanol, geometry = geometry),
    stat = "sf_coordinates",
    size = 2.5
    ) + 
  
  # now let's add in our customized graticules too. 
  scale_x_continuous(breaks = brks$x) +
  scale_y_continuous(breaks = brks$y) + 
  theme_nautical() + 
  
  labs(
    x = NULL, y = NULL,
    title = 'geom_text_repel(places)', 
    subtitle = '+ land + places'
    )

m
```

Because the package is attached, we can just start using the data — it's currently held as a promise.
In other words, we don't need to call `data(object)` on the data sets; we can use them directly (for example, by calling `head(object)`).
We'll use this direct approach for the remainder of the vignette.

### route

```{r show the routes, warning=F}
head(route)

route <- left_join(
  route,
  st_drop_geometry(places),
  by = c('destination' = 'location_english')
  ) |>
  relocate(geometry, .after = last_col())

```

The `route` object itself is pretty minimal, but relevant attributes can be brought in by joining it to `places`.
```{r route head, warning=F}
head(route)
```

While some of the data for the start and end of the trip is included — such as entries near Santa Barbara and the departure from and return to Monterey — most of it focuses on the Gulf of California.
Basically, including too much data would make the package too cumbersome to fit on CRAN.
In my mind, the maximum useful map area reaches San Diego in the north.

```{r}
date_scale <- as.Date(quantile(as.numeric(route$date_arrive), na.rm = T, probs = c(0.1, 0.5, 0.9)))
dates <- c(
  expression(paste("March ", 16^th)), 
  expression(paste("Mar. ", 28^th)), 
  expression(paste("April ", 11^th))
  )
```

```{r subset bounding box, warning = F}
bb <- st_bbox(
  c(xmin = -114, xmax = -108.5, ymin = 22.5, ymax = 30.5),
  crs = st_crs(4326)
)

# we have to crop the places data set or ggrepel will move places outside the
# coord_sf down into the plot anyways

places <- st_crop(places, bb)

m <- ggplot()  +
  geom_sf(data = land, fill = '#F8F6F0') + 
  theme_nautical() +
  geom_sf(data = route, aes(color = date_arrive)) +  
 # scale_color_date() +  # alteratively just use this and default labels. 
  scale_color_continuous('Date',
      breaks = date_scale,
      labels = dates
      )  + 
  coord_sf(xlim = c(bb[1], bb[3]), ylim = c(bb[2], bb[4])) + 
  
  labs(
    title = 'route', 
    subtitle = '+ land')

  
m
```

Now let's add some topography to make the land seem more natural.
We'll also ignore the administrative borders.

Note that we're going back to the drawing board to control the order in which layers are added to the map.
We'll still overwrite the variable `m`.

```{r add topography}

m <- ggplot() + 
  geom_sf(data = land, fill = '#F8F6F0') + 
  geom_sf(data = topography, lwd = 0.1) + 
  geom_sf(data = route, aes(color = date_arrive)) +
  scale_color_date('Date') + 
  coord_sf(xlim = c(bb[1], bb[3]), ylim = c(bb[2], bb[4])) + 

  # we are going to shift to adding 'backing' to the labels this makes them easier to read
  ggrepel::geom_label_repel(
    data = places,
    aes(label = location_espanol, geometry = geometry),
    stat = "sf_coordinates",
    alpha = 0.7, # make the backing more transparent
    label.size = NA, # remove the backing borders
    label.padding = 0.1, # reduce space between label borders and text
    size = 2.5 # make the font smaller. 
    ) + 
    scale_color_continuous('Date',
      breaks = date_scale,
      labels = dates
      ) + 
  
  # now let's add in our customized graticules too. 
  scale_x_continuous(breaks = brks$x) +
  scale_y_continuous(breaks = brks$y) + 
  theme_nautical() + 
  labs(
    x = NULL, y = NULL,
    title = 'elevation', 
    subtitle = '+ route + places') 

m
```

We can plot the bathymetry data like this.

```{r}
ggplot() + 
  geom_sf(data = land, fill = '#F8F6F0') + 
  geom_sf(data = bathymetry, aes(color = elevation), lwd = 0.4)  +
  theme_nautical() + 
  coord_sf(xlim = c(bb[1], bb[3]), ylim = c(bb[2], bb[4])) + 
  labs(title = 'bathymetry', subtitle = '+ land')
```
And obviously we could rename it to something like depth :)

Alternatively, the same scale can be used for topography and bathymetry together, as shown below. I'll use a divergent scale, which makes sense to me here.
Another *very* cool interpretation would be a continuous scale that counts everything from 0 at 4,300 feet and adds the difference to the `topography` data set.
Or — much to my liking — we could convert the bathymetry polylines to polygons and use them to color the whole ocean, with darker areas as deeper hues of blue.

```{r and bathymetry}
head(bathymetry)

ggplot() + 
  geom_sf(data = land, fill = '#F8F6F0') + 
  geom_sf(data = bathymetry, aes(color = elevation), lwd = 0.4) + 
  geom_sf(data = topography, aes(color = elevation), lwd = 0.4) + 
  scale_color_distiller('Elevation', palette = "Spectral") +
  
  ggnewscale::new_scale_color() + 

  geom_sf(data = places) +  
  geom_sf(data = route, aes(color = date_arrive)) +
  labs(title = 'topography', subtitle = '+ land + bathymetry')+
  scale_color_continuous('Date',
      breaks = date_scale,
      labels = dates
      ) + 
  theme_nautical() + 
  coord_sf(xlim = c(bb[1], bb[3]), ylim = c(bb[2], bb[4]))

```


## Tangential data

While some of the earlier data sets are loosely related to the book, these next two aren't related at all — but they can be useful for cartography.

A simple landcover classification data set is available as `landcover`.
We also include some quick colors to help with mapping these classes.

```{r}
ggplot() + 
  geom_sf(data = landcover, aes(fill = class), color = NA) + 
  scale_fill_manual('Class', values = lc_pal, breaks = names(lc_pal[c(1:8, 9)])) + 
  theme_nautical() + 
  labs(title = 'landcover', subtitle = '+ lc_pal') + 
  guides(fill = guide_legend(nrow = 3)) + 
  coord_sf(xlim = c(bb[1], bb[3]), ylim = c(bb[2], bb[4]))
```

Information on protected areas of Mexico is also included.

```{r Protected areas, warning = F}
protected <- st_crop(protected, bb)

ggplot() + 
  geom_sf(data = land, fill = '#F8F6F0') + 
  geom_sf(
    data = protected, 
    aes(fill = reserve_type), 
    alpha = 0.4) + 
  scale_fill_manual('Reserve', values = c('Terrestrial' = '#417B5A', 'Marine' = '#DA7635')) + 
  ggrepel::geom_text_repel(
    data = protected,
    aes(label = name, geometry = geometry),
    stat = "sf_coordinates",
    size = 2.5
    ) + 
  labs(x = NULL, y = NULL, title = 'protected', subtitle = '+ land') + 

  theme_nautical()  + 
  coord_sf(xlim = c(bb[1], bb[3]), ylim = c(bb[2], bb[4]))
```

# Putting it all together

We can make an OK map using some of the details below.

```{r}
ggplot() + 
  
  # the stage
  geom_sf(data = landcover, aes(fill = class)) + 
  scale_fill_manual('Class', values = lc_pal) + 
  geom_sf(data = land, fill = '#F8F6F0', alpha = 0.5) +  # helps to dull the landcover for this map. 
  guides(fill="none") + 

  geom_sf(data = bathymetry, aes(color = elevation), lwd = 0.25) + 
  geom_sf(data = topography,  aes(color = elevation), lwd = 0.25, color = 'black') + 
#  scale_color_distiller('Elevation', palette = "Spectral") +
  guides(color='none') + 
  ggnewscale::new_scale_colour() + 
  
  # the story 
  geom_sf(data = route, aes(color = date_arrive)) +
    scale_color_continuous('Date',
      breaks = date_scale,
      labels = dates
      ) + 

  geom_sf(data = places, shape = 5, size = 1.5, color = '#222823') +   
  ggrepel::geom_label_repel(
    data = places,
    aes(label = location_espanol, geometry = geometry),
    stat = "sf_coordinates",
    alpha = 0.7, # make the backing more transparent
    label.size = NA, # remove the backing borders
    label.padding = 0.1, # reduce space between label borders and text
    size = 2.5,# make the font smaller. 
    fill = '#F4F7F5'
    ) +
  
  # the ambiance.  
  coord_sf(xlim = c(bb[1], bb[3]), ylim = c(bb[2], bb[4]), crs = 4326) + 
  annotation_scale(bar_cols = c('#222823', '#F4F7F5'))  + 
  annotation_north_arrow(which_north = "true", style = north_arrow_nautical) + 
  theme_nautical() +
  labs(title = 'A map', subtitle = 'Ricketts <3 giardiniera', x = NULL, y = NULL)

```

```{r warning = FALSE}
rm(graticNo, brks, theme_nautical, route, places, bb, landcover, lc_pal, land,
   protected, bathymetry, topography)
```


