---
title: "Exploring the data sets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exploring the data sets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction and quick start 

flyer is available on github, and should go to CRAN sometime. 
It can be installed via either of  the following commands. 

```{r install package, eval = F}
install.packages('devtools')
devtools::install_github('sagesteppe/flyer')

# remotes is very similar and a good alternative for this use case.
install.packages('remotes') 
remotes::install_github('sagesteppe/flyer')
```

To explore the data we will load a couple packages for interacting with the data (sf, dplyr), and a whole slew of packages for plotting the data using ggplot2 (ggnewscale, ggrepel). 
It might seem onerous to be installing all of these packages, but I bet once you see what they do you'll be quite excited to have them. 

```{r setup}
library(flyer)
library(dplyr) # for general data handling
library(sf) # for spatial data

library(ggplot2) # all for plotting the data 
library(ggnewscale) # for mapping multiple variables to an aesthetic. 
library(ggrepel) # for text based labels which move to minimize overlaps. 
library(ggspatial) # compasses and scale bars. 
```

We'll modify the number of graticules right off the bat, note that the `pretty` function does not always return the requested `n`, so... 

```{r define a carto ggplot theme}

graticNo <- function(polygon, nx, ny){
  
  bb <- round(st_bbox(polygon), 1)
  if(all(missing(nx) & missing(ny))) {nx <- 5;ny <- 5} else {
    if(all(missing(nx) & ! missing(ny))) {nx <- ny} else {
      if(all(missing(ny) & ! missing(nx))) {ny <- nx}
    } 
  } 

  xbreaks <- pretty(seq(bb[1], bb[3], length.out = nx), nx)
  ybreaks <- pretty(seq(bb[2], bb[4], length.out = ny), ny)
  
  return(
    list(
      x = xbreaks, y = ybreaks
    )
  )
}

brks <- graticNo(polygon = places, nx = 4, ny = 5)

```

# the data sets 

The places visited by the collectors can be loaded using `places`, and the route they took can be loaded via `routes`.
These are really the whole point of the package!
And spoiler alert... 
are very simple!

```{r plot the places}
data(places)
head(places)

m <- ggplot() + 
  geom_sf(data = places)
```

The places seem like they will be better treated as text labels - we can apply them with `ggrepel::geom_text_repel` which will move them to avoid conflicts with other plot elements. 

```{r plot places as text, warning=F}
m <- ggplot() +
  
  ggrepel::geom_text_repel(
    data = places,
    aes(label = location_espanol, geometry = geometry),
    stat = "sf_coordinates",
    size = 3
    ) + 
  
  # now let's add in our customized graticules too. 
  scale_x_continuous('Longitude', breaks = brks$x) +
  scale_y_continuous('Latitude', breaks = brks$y) + 
  theme_minimal()

m
```

Also, because the package is attached we can also just start using the data, which is currently held as a promise. 
In other words we don't need to call `data(object)` on the data sets, we can just use them directly - such as viewing them by calling `head(object)`. 
We are going to use this direct approach for the remainder of the vignette. 

```{r show the routes, warning=F}
head(route)

route <- left_join(
  route,
  st_drop_geometry(places),
  by = c('destination' = 'location_english')
  ) |>
  relocate(geometry, .after = last_col())

```

the route object itself is pretty minimal, but relevant attributes can be brought in by joining it to places. 
```{r route head, warning=F}
head(route)
```

By default ggplot2 does not seem to order date variables the way I would. 
We can convert the dates to factor to enforce data to be plotted in actual order. 

```{r, warning=F}
m + 
  geom_sf(data = route, aes(color = date_arrive))
```

Obviously we need some context for these data. 
Other wise they just... look the way they do above. 
We can read in some polygons which depict land from `Natural Earth`, from the `rnaturalearth` package; note that I love this packages functionality, even if I get real forgetful of their API calls (theme argument?). 
Anyways, we have duped the natural earth land surfaces into this package - but there are ton's of other products you may want to use from there, check out the [R Open Sci page for more info](https://docs.ropensci.org/rnaturalearth/articles/rnaturalearth.html). 

```{r add land}
head(land)

ggplot() + 
  geom_sf(data = land)
```

For playing around with the data today, I don't want the different countries to be drawn separately so we can union them.  

```{r union land}
land <- st_union(land)

ggplot() + 
  geom_sf(data = land) + 
  theme_minimal()
```

While some of the data for the start end of the trip is included, such as an entry when they are near Santa Barbara and that they started and returned to Monterey, most all data focuses on the Gulf of Mexico. 
Basically, too much data would make the package too cumbersome to fit on CRAN. 

```{r subset bounding box}

bb <- st_bbox(
  c(xmin = -114, xmax = -108.5, ymin = 22.5, ymax = 30.5),
  crs = st_crs(4326)
)

# we have to crop the places data set or ggrepel will move places outside the
# coord_sf down into the plot anyways

places <- st_crop(places, bb)

m <- ggplot()  +
  geom_sf(data = land) + 
  theme_minimal() +
  coord_sf(xlim = c(bb[1], bb[3]), ylim = c(bb[2], bb[4]))

m
```

Now let's add some topography to make the land seem more natural. 
We'll also ignore the administrative borders. 

Note that we are going to go back to the drawing board to control the order which layers are added to the map.
We'll still overwrite the variable m. 

```{r add topography}

plot(topography)

m <- ggplot() + 
  geom_sf(data = land) + 
  geom_sf(data = topography, lwd = 0.1) + 
  geom_sf(data = route, aes(color = date_arrive)) +
  coord_sf(xlim = c(bb[1], bb[3]), ylim = c(bb[2], bb[4])) + 

  
  ggrepel::geom_text_repel(
    data = places,
    aes(label = location_espanol, geometry = geometry),
    stat = "sf_coordinates",
    size = 2.5
    ) + 
  
  # now let's add in our customized graticules too. 
  scale_x_continuous('Longitude', breaks = brks$x) +
  scale_y_continuous('Latitude', breaks = brks$y) + 
  theme_linedraw()

m
```

We can plot the bathymetry data like this. 

```{r}

ggplot() + 
  geom_sf(data = land, fill = '#EFDD8D') + 
  geom_sf(data = bathymetry, aes(color = elevation), lwd = 0.4)  +
  theme_minimal() + 
  coord_sf(xlim = c(bb[1], bb[3]), ylim = c(bb[2], bb[4]))
```
And obviously we can rename it something like depth... :)

On the other hand, the same scale can be used for topography and bathymetry together like this... I will use a divergent scale which makes sense to me... 
I could see another *very* cool interpretation of using a continuous and counting everything from 0 at 4300 feet and adding the difference to the 'topography' data set. 
Or, and much to my liking, we can convert the polylines of bathymetry to polygons, and use them to color the whole ocean!
With darker areas being deeper hues of blue. 

```{r and bathymetry}
head(bathymetry)
plot(bathymetry)

ggplot() + 
  geom_sf(data = land, fill = '#EFDD8D') + 
  geom_sf(data = bathymetry, aes(color = elevation), lwd = 0.4) + 
  geom_sf(data = topography, aes(color = elevation), lwd = 0.4) + 
  scale_color_distiller(palette = "Spectral") +
  
  ggnewscale::new_scale_color() + 

  geom_sf(data = places) +  
  geom_sf(data = route, aes(color = date_arrive)) +
  theme_minimal() + 
  coord_sf(xlim = c(bb[1], bb[3]), ylim = c(bb[2], bb[4]))

```


## tangential data 

While some of the last data sets are loosely related to the book, these two are... not at all related, but can be useful for cartography. 

A simple landcover classification data set is available as `landcover`. 
We also include some quick colors to help with mapping these classes. 

```{r}

ggplot() + 
  geom_sf(data = landcover, aes(fill = class)) + 
  scale_fill_manual(values = lc_pal) + 
  theme_minimal()

```

Additionally, information on protected areas of Mexico are also included. 

```{r}
ggplot() + 
  geom_sf(data = land) + 
  geom_sf(data = protected, aes(fill = reserve_type)) + 
  theme_minimal()

```

# putting it all together

We can make an OK map using some of the details below. 

```{r}

ggplot() + 
  geom_sf(data = land, fill = '#EFDD8D') + 
  geom_sf(data = bathymetry, aes(color = elevation), lwd = 0.1) + 
  geom_sf(data = topography,  aes(color = elevation), lwd = 0.1) + 
  geom_sf(data = places) +  
 # geom_sf(data = route, aes(color = date_arrive)) +
  coord_sf(xlim = c(bb[1], bb[3]), ylim = c(bb[2], bb[4]), crs = 4326) + 
  annotation_scale()  + 
  annotation_north_arrow(which_north = "true", style = north_arrow_minimal)

  north_arrow_minimal()
```

