---
title: "PenguinClassicsMap"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PenguinClassicsMap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r install package, eval = F}
devtools::install_github('sagesteppe/flyer')
```

The Penguin Classics map has many wonderful attributes worth emulating. 

```{r setup}
library(flyer)
library(sf)
library(dplyr)
library(ggplot2)
```

The places visited by the collectors can be loaded using `places`, and the route they took (which I made up) can be loaded via `routes``. 

```{r plot the places}
data(places)
head(places)

ggplot() + 
  geom_sf(data = places)
```

Because the package is attached we can also just start using the data, which is currently held as a promise. 

```{r show the routes}

data(route)

route <- left_join(
  route,
  st_drop_geometry(places),
  by = c('destination' = 'location_english')
  )

ggplot() + 
  geom_sf(data = places) + 
  geom_sf(data = route, aes(color = date_arrive)) 

```

Obviously we need some context for these data. 
Other wise they just... look the way they do above. 
We can read in some polygons which depict land from `Natural Earth`, from the `rnaturalearth` package; note that I love this packages functionality, even if I get real forgetful of their API calls (theme argument?)

```{r add land}
data(land)
head(land)

ggplot() + 
  geom_sf(data = land) + 
  geom_sf(data = places) + 
  geom_sf(data = route, aes(color = date_arrive)) 

```

While some of the data for the start end of the trip is included, such as an entry when they are near Santa Barbara and that they started and returned to Monterey, most all data focuses on the Gulf of Mexico. 
Basically, too much data would make the package too cumbersome to fit on CRAN. 


```{r}

bb <- st_bbox(
  c(xmin = -114, xmax = -108.5, ymin = 22.5, ymax = 30.5),
  crs = st_crs(4326)
)

ggplot() + 
  geom_sf(data = land) + 
  geom_sf(data = places) + 
  geom_sf(data = route, aes(color = date_arrive)) +
  coord_sf(xlim = c(bb[1], bb[3]), ylim = c(bb[2], bb[4]))


```


Now let's add some topography to make the land seem more natural. 
We'll also ignore the administrative borders. 

```{r add topography}

land <- st_union(land)
plot(topography)

ggplot() + 
  geom_sf(data = land) + 
  geom_sf(data = topography, lwd = 0.1) + 
  geom_sf(data = places) + 
  geom_sf(data = route, aes(color = date_arrive)) +
  coord_sf(xlim = c(bb[1], bb[3]), ylim = c(bb[2], bb[4]))

```


```{r and bathymetry}
head(bathymetry)
plot(bathymetry)

ggplot() + 
  geom_sf(data = land, fill = '#EFDD8D') + 
  geom_sf(data = bathymetry, lwd = 0.1) + 
  geom_sf(data = topography, lwd = 0.1) + 
  geom_sf(data = places) +  
  geom_sf(data = route, aes(color = date_arrive)) +
  coord_sf(xlim = c(bb[1], bb[3]), ylim = c(bb[2], bb[4]))

```
