---
title: "PenguinClassicsMap"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PenguinClassicsMap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r install package, eval = F}
devtools::install_github('sagesteppe/flyer')
```

The Penguin Classics map has many wonderful attributes worth emulating. 

```{r setup}
library(flyer)
library(sf)
library(tidyverse)
```

The places visited by the collectors can be loaded using `places`, and the route they took (which I made up) can be loaded via `routes``. 

```{r}
data(places)
head(places)

ggplot() + 
  geom_sf(data = places)
```

Because the package is attached we can also just start using the data, which is currently held as a promise. 

```{r}
data(route)
plot(route)

route <- left_join(
  route,
  st_drop_geometry(places),
  by = c('Destination' = 'location.english')
  )

ggplot() + 
  geom_sf(data = places) + 
  geom_sf(data = route, aes(color = date_arrive)) 

```

Obviously we need some context for these data. 
Other wise they just... look the way they do above. 
We can read in some polygons which depict land from `Natural Earth`, from the `rnaturalearth` package; note that I love this packages functionality, even if I get real forgetful of their API calls (theme argument?)



